Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-03-28 paper_jasa.tex
some small changes to reduce size

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables statistical significance testing of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics are developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. \blue{Results suggest that the lineup protocol performs comparably with the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated. Surprisingly, visual tests have higher power than the conventional tests when the effect size is large. And, interestingly, there may be some super-visual individuals who yield better performance and power than the conventional test even in the most difficult tasks.}

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, \blue{and unlikely ever to be any numerical test}. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done based on a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

\blue{An illustration of the lineup protocol in relation to conventional hypothesis testing is presented in Table \ref{tbl:compare}.} Both methods start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

\begin{table*}[hbtp]
\caption{Comparison of visual inference with conventional inference.}
\centering 
\begin{tabular}{llll} 
\hline \hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

%Visual tests are non-parametric tests of the form $H_0$: the data plot is visually not distinguishable from a null plot vs $H_1$: the data plot is visually distinguishable from a plot generated consistently with $H_0$.

%Power of a plot and $p$-value have been derived in Buja and Wickham as ...

%for this paper, we want to place the parametric inference of a regression setting into the framework of visual testing:

%This section outlines the concepts of visual inference in comparison to the procedures of conventional statistical inference. Table \ref{tbl:compare} (derived from \citet{buja:2009}) gives a summarized overview of this comparison.

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that maps the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T(y_0)$ which would be possible values of $T$ if $\theta \in \Theta$.  %Similarly, for any test statistic, we have a rejection region $R$ and, complementary to that, $R^c$,  that partitions the space of the test statistic, which is usually a subspace of the real numbers. 
For visual inference, unlike in the conventional hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the actual data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

@@

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

\begin{dfn}\label{dfn:lplot}
A \textbf{lineup} is a layout of $m$ randomly placed visual statistics, consisting of 
\begin{itemize}\itemsep-3pt
\item $m-1$ statistics, $T(y_0)$, simulated from the model specified by $H_0$  (null plots) and 
\item the test statistic, $T(y)$, produced by plotting the actual data, possibly arising from $H_1$.
\end{itemize}
\end{dfn}

\noindent The $(m-1)$ null plots are % can be 
members of the sampling distribution of the test statistic assuming that the null hypothesis is true. If $H_1$ is true, we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots. A careful visual inspection of the lineup by %\strikethrough{an} 
independent observers follows;  observers are asked to point out the plot most different from the lineup. If the test statistic is identified in the lineup, this is considered as evidence against the null hypothesis. 
%If it is not identified the conclusion is to not reject the null hypothesis.

This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}_{K, 1/m}$,  therefore:

@@

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision. Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of  size $m$ and significance level $\alpha$, is defined to  
\begin{itemize}\itemsep-3pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}|H_0) \le \alpha$. 
\end{dfn}

\noindent
Associated with any test is the risk of  Type I or II errors, which for visual inference are defined as follows:

\begin{dfn}\label{dfn:error}
The \textbf{Type I error} associated with visual test $V_{\theta}$ is the probability of rejecting $H_0$ when it is true; the probability for that is $P(X \ge x_{\alpha})$, which is controlled by $\alpha$.\\ \noindent
The \textbf{Type II error} is the probability of failing to identify the actual data plot, when $H_0$ is not true, $P( X <  x_{\alpha})$.
\end{dfn}

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual $i$ picks out the actual data plot from the lineup. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot from the lineup.

@@

\noindent An important difference between conventional and visual testing is that lineups will depend on observers' evaluation. Thus $X$, the number of observers who identify the actual data plot from the lineup, affects the estimation of power  and the power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
Here $F_{X, \theta}$ is the distribution of $X$ and $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. Note that the distribution  $F_X$ depends on which hypothesis is true:
under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
Power_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
%where $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
If the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's skill to identify the actual data plot will require that an individual evaluates multiple lineups.

Power is an important consideration in deciding which test to use for solving a problem. %In conventional testing, new test procedures are examined for their power under a variety of scenarios. 
Here we use it to compare the performance of the visual test with the conventional test, but in practice for visual inference it will mostly be important in deciding on the choice of plot to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test.

@@

%Let $X_{\ell i} \sim Binom_{1, p_{\ell i}}$ be the random variable, with $X_{ \ell i}= 1$, if  subject $i$ correctly identifies the actual data plot from lineup $\ell$, and 0 otherwise.
The model can be  fit as:% $p_{ \ell i}$ with a logistic regression with subject-specific random effects as:
\begin{equation} \label{eqn:mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

%\red{This model is  described in more detail in equation \eqref{eqn:mixed}.}

%\noindent For any single observer, the power of a visual test has a lower limit of $\frac1m$ since the probability that an observer will randomly pick the test statistic under $H_0$ is $\frac1m$. When $K$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (K,1/m)$  under $H_0$. This leads us to obtain the probability 
%\begin{equation}\label{eqn:pvalue}
%Pr(U \ge u)= \sum_{k \ge u}^K {{K \choose k} \left(\frac{1}{m}\right)^k\left(1-\frac 1m\right)^{(K-k)}}
%\end{equation}
%where $u$ is the observed number of successful evaluations. We use this probability to  define the $p$-value for visual test $V_{\theta}$ as follows:
%\begin{dfn} \label{dfn:pval}
%The $p$-value associated with the decision of a visual test $V_{\theta}$ is given \blue{as the probability to observe a value at least as extreme as the test statistic $u$, 
%where $u$ is the number of successful evaluations of the lineup, which leads us to a $p$-value of $P(U \ge u)$ as given in eqn (\ref{eqn:pvalue}). Since $u$ is a discrete value with boundaries 0 and $K$,
% we can also re-formulate the $p$-value in terms of upper and lower boundaries:
% \begin{equation*}
% \begin{array}{rcll}
% P(U \ge u+1) \le &p\text{-value}& (\le 1), & \text{if }u < K, \text{and in particular for } u=0\\
%( 0 \le) &p\text{-value}& \le P(U \ge u-1), & \text{if }u > 1, \text{and in particular for } u=K.\\ 
% \end{array}
% \end{equation*}
% }

%\red{HH: Tried to re-write it to answer the question below.
%Why is the $p$ value not just $Pr(U \ge u)$ ? (We need to modify this a bit with respect to the rejection region $R$, but there should not be a difference in terms of whether we reject or accept. }
%
%    \begin{equation*}
%        \begin{cases} 
%              p\text{-value} \le Pr(U \ge u)    & \text{when reject $H_0$,} \\
%              p\text{-value} > 1-Pr(U \ge u)  & \text{when fail to reject $H_0$}
%        \end{cases}
%    \end{equation*}
%\end{dfn}

%\red{HH: the following paragraph needs some adjustment in terms of t and $\theta$ - could you try to take of that, Mahbub?}\\
%\red{For the following paragraph, we wanted to get rid of the absolute value of $t$ and instead look at $t \in R$ and $t \in R^c$.}

%\noindent
\subsection{Lineup size, $m$}

@@

\begin{lemma}~\label{lemma}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the conventional test statistic. Suppose the associated test statistic in a conventional test is observed as  $t_{obs}$ with $p$-value $p_D$.

@@

\begin{proof} The proof and details of the lemma is attached in the supplementary documents. 
\end{proof}

@@

The conventional test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis, if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom.  It forms the benchmark upon which we evaluate the visual test. To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the conventional $p$-value as details given in Lemma \ref{lemma}. \blue {Thus the expected probability to reject by a single observer ($K=1$) in this scenario is

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$ gets larger, and when $\beta$ gets larger.  } \hh{Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. This dichotomy of visual power can also be seen in more detail in figure \ref{fig:power_loess_effect}. }

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  \blue {Figure \ref{fig:test_category} is an example of one of these lineups. It was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. The actual data plot location is ($4^2-1$). For this lineup, 15 out of 16 observers picked the actual data plot. }

@@

The number of evaluations required for each lineup to provide reasonable estimates of the \blue{ proportion correct ($\hat p$) is determined by the variance of the number of correct evaluations}. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) \blue{we consider $\gamma =p$}.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

%For a given proportion $p$ we want to have margin of error (ME) to be 0.1. Thus we have $ME =1.96 \sqrt{ \frac 1 n p(1-p)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_p \geq \frac{p(1-p)}{(0.05/1.96)^2}.$$ Now from the theoretical (UMP) power curve for each parameter combinations shown in Table \ref{tbl:experiment_params} we know the power(say $p$) and thus we can estimate the number of evaluations required for that specific combinations of parameter. We collect data such that each of the parameter combinations in the Table \ref{tbl:experiment_params} has approximately the estimated sample size that we calculated here.

%\green{Did you in fact bin like this???}\red{For each parameter combination a set of 1000 actual data are generated and their associated p-values were computed. This produces an empirical distribution of p-value under alternative hypothesis. Finally, three data sets that have 16th, 50th and 84th quantile of the p-values were taken. Similar procedure was used to generate 19 null plots based on the empirical distribution of p-value under null hypothesis}. For added control, to ensure a signal in the simulated actual data a blocking structure was used to filter data sets. A 1000 sets were generated for each parameter combination and the conventional $t$-statistic and $p$-value associated with $H_0: \beta_2=0$ were calculated. The 3 replicates were drawn from each of three blocks of $p$-values: (0.0-$q_{33}$), ($q_{33}$-$q_{66}$), ($q_{66}$-1) where $q_i$ is the $i$th percentile in the distribution of the $p$-values. Additional control was applied to the 19 null plots. Because the distribution of these $p$-values should follow a Uniform(0,1) distribution, data sets were binned on this range by $p$-value, and a data set was randomly selected from each bin.

@@

This experiment is very similar to the previous one, except that there is a single continuous covariate and no second covariate (Equation \ref{multi} with $p=1$), following the test in Table \ref{tbl:stat_multiple}, case 2. Data is simulated with two sample sizes ($n=100, 300$), two standard deviations of the error ($\sigma=5, 12$), a variety of slopes ($\beta$), as given in Table  \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ and values for $X_1$ are simulated from $N(0,1)$. \blue{For each combination of parameters, at least three different actual data sets are produced, yielding a total of 70 lineups. }

%For some parameter combinations we have more than three replications, because additional lineups were produced for values of the parameters that exhibited large variation in the initial data.

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. A new method, called the lineup protocol, enables statistical significance testing of visual findings, which bridges a big gulf between exploratory and inferential statistics.  In this paper inferential methods for statistical graphics are developed further by refining the terminology of visual inference, and framing the lineup protocol in a context that allows direct comparison with conventional tests, in scenarios when a conventional test exists. This framework is used to compare the performance of the lineup protocol against conventional statistical testing in the scenario of fitting linear models. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol performs comparably with the conventional tests, and expectedly out-performs them when data is contaminated, a scenario where assumptions required for performing a conventional test are violated. Surprisingly, visual tests have higher power than the conventional tests when the effect size is large. And, interestingly, there may be some super-visual individuals who yield better performance and power than the conventional test even in the most difficult tasks.

@@

The protocol has only been informally tested until now. In the work described in this paper, the lineup protocol is compared head to head with the equivalent conventional test. Specifically, the lineup is examined in the context of a linear model setting, where we are determining the importance of including a variable in the model. This is not the envisioned environment for the use of the lineup -- actually it is likely the worst case scenario for visual inference. The intended use of lineups is where there is no existing test, and unlikely ever to be any numerical test. The thought is though, that the conventional setting provides a benchmark for how well the lineup protocol works under controlled conditions, and will provide some assurance that they will work in scenarios where there is no benchmark. Testing is done based on a human-subjects experiment using Amazon's Mechanical Turk \citep{turk}, using simulation to provide controlled conditions for assessing lineups. The results are compared with those of the conventional test.

@@

An illustration of the lineup protocol in relation to conventional hypothesis testing is presented in Table \ref{tbl:compare}. Both methods start from the same place, the same set of hypotheses. The conventional test statistic is the $t$-statistic, where the parameter estimate is divided by its standard error. In the lineup protocol, the test statistic is a plot of the data. Here, side-by-side boxplots are used, because the variable of interest is categorical and takes just two values. In conventional hypothesis testing the value of the test statistic is compared with all possible values of the sampling distribution, the distribution of the statistic if the null hypothesis is true. If it is extreme on this scale then the null hypothesis is rejected. In contrast in visual inference, the plot of the data is compared with plots of a finite number of samples drawn from the null distribution. If the actual data plot is selected as the most different, then this results in rejection of the null hypothesis.

\begin{table*}[hbtp]
\caption{Comparison of visual inference with conventional inference.}
\centering 
\begin{tabular}{llll} 
\hline % \hline
  & Conventional Inference &  Lineup Protocol \\ %[0.5ex] % inserts table %heading 
\hline
  Hypothesis & $H_0: \beta=0$ vs $H_1: \beta > 0$& $H_0: \beta=0$ vs $H_1: \beta > 0$\\
 & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} & \begin{minipage}[h]{2.5cm} \begin{center} \scalebox{0.2}{\includegraphics{down_arrow.pdf}} \end{center} \end{minipage} \\

@@

In general, we define $\theta$ to be a population parameter of interest, with $\theta \in \Theta$, the parameter space. Any null hypothesis $H_0$ then partitions the parameter space into $\Theta_0$ and $\Theta_0^c$, with $H_0: \theta \in \Theta_0$ versus $H_1: \theta \in \Theta_0^c$. A test statistic, $T(y)$, is a function that maps the sample into a numerical summary, that can be used to test the null hypothesis. The hypothesis test maps the test statistic into \{0, 1\}, based on whether $T(y)$ falls into the acceptance region, or the rejection region, respectively. $T(y)$ is assessed relative to null values of this statistic $T(y_0)$, the possible values of $T$ if $\theta \in \Theta$.

For visual inference, unlike in the conventional hypothesis test, the statistic is not a single value, but a graphical representation of the data chosen to display the strength of the parameter of interest, $\theta$. When the alternative hypothesis is true, it is expected that the plot of the actual data, the test statistic, will have visible feature(s) consistent with $\theta \in \Theta_0^c$, and that visual artifacts will not distinguish the test statistic as different when $H_1$ is not true. We will call a plot with this property a {\it visual statistic} for $\theta$. More formally,

@@

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type and specification of aesthetics, necessary for complete reproducibility. The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

\begin{dfn}\label{dfn:lplot}
A \textbf{lineup} is a layout of $m$ randomly placed visual statistics, consisting of 
\begin{itemize}\itemsep-10pt
\item $m-1$ statistics, $T(y_0)$, simulated from the model specified by $H_0$  (null plots) and 
\item the test statistic, $T(y)$, produced by plotting the actual data, possibly arising from $H_1$.
\end{itemize}
\end{dfn}

\noindent The $(m-1)$ null plots are % can be 
members of the sampling distribution of the test statistic assuming that the null hypothesis is true. If $H_1$ is true, we expect this to be reflected as a feature in the test statistic, i.e. the plot of the data, that makes it visually distinguishable from the null plots. A careful visual inspection of the lineup by %\strikethrough{an} 
independent observers follows;  observers are asked to point out the plot most different from the lineup. If the test statistic is identified in the lineup, this is considered as evidence against the null hypothesis.  This leads us to a definition for the $p$-value of a lineup: under the null hypothesis, each observer has a $1/m$ chance of picking the test statistic from the lineup.  For $K$ independent observers, let $X$ be  the number of observers picking the test statistic from the lineup. Under the null hypothesis $X \sim \text{Binom}_{K, 1/m}$,  therefore:

@@

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. By increasing either $m$ or $K$, the value at a higher precision can be determined. Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

@@

\begin{dfn}\label{dfn:test}
The \textbf{visual test}, $V_{\theta}$  of  size $m$ and significance level $\alpha$, is defined to  
\begin{itemize}\itemsep-10pt
\item \textbf{Reject} $H_0$ if out of $K$ observers at least $x_{\alpha}$ correctly identify the actual data plot, and
\item \textbf{Fail to reject} $H_0$  otherwise. 
\end{itemize}
where $x_{\alpha}$ is such that $P(X \ge x_{\alpha}|H_0) \le \alpha$. 
\end{dfn}

\noindent
Associated with any test there is the risk of  Type I or II errors, which for visual inference are defined as follows:

\begin{dfn}\label{dfn:error}
The \textbf{Type I error} associated with visual test $V_{\theta}$ is the probability of rejecting $H_0$ when it is true; the probability for that is $P(X \ge x_{\alpha})$, which is controlled by $\alpha$. The \textbf{Type II error} is the probability of failing to identify the actual data plot, when $H_0$ is not true, $P( X <  x_{\alpha})$.
\end{dfn}

\noindent Because $X$ takes only discrete values we can not always control exactly for $\alpha$.  
For example, when there is only one observer, $1/m$ is the minimal value at which we can set $\alpha$. It can be set to be smaller, even arbitrarily small, by increasing $K$, the number of observers. 
Type II error is harder to calculate, as is usually the case. In visual inference, individual abilities need to be incorporated to calculate Type II error. %This is not necessary for Type I error, because it is simply the chance that the observer picks the actual data plot, by chance. 
Here, we need to estimate the probability that an observer sees the actual data plot as different, when it really is different. This involves understanding the individual's visual skills. Thus, let $X_i$ be a binary random variable with $X_i = 1$, if individual $i \ (=1, \dots , K)$ identifies the actual data plot from the lineup, and $X_i = 0$ otherwise. Let $p_i$ be the probability that individual $i$ picks out the actual data plot. If all individuals have the same ability, with the probability, $p$, for picking out the actual data plot, then $X = \sum_i X_i$ has distribution $\text{Binom}_{K, p}$, and we can estimate $p$ by $\hat{p} = x/K$, where $x$ is the number of observers (out of $K$), who pick out the actual data plot.

@@

\noindent An important difference between conventional and visual testing is that lineups will depend on observers' evaluation. Thus $X$, the number of observers who identify the actual data plot from the lineup, affects the estimation of power  and the power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1).
\] 
Here $F_{X, \theta}$ is the distribution of $X$ and $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. Note that the distribution  $F_X$ depends on which hypothesis is true:
under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
Power_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1).
\]
If the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's skill to identify the actual data plot will require that an individual evaluates multiple lineups.

Power is an important consideration in deciding which test to use for solving a problem. Here we use it to compare the performance of the visual test with the conventional test, but in practice for visual inference it will mostly be important in choosing plots to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test.

@@

The model can be  fit as:
\begin{equation} \label{eqn:mixed}
g( p_{\ell i} )= W_{\ell i} \delta +  Z_{\ell i}  \tau_{\ell i},  
\end{equation}

@@

\subsection{Lineup size, $m$}

@@

\begin{lemma}~\label{lemma}
Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the conventional test statistic. Suppose the associated test statistic is observed as  $t_{obs}$ with $p$-value $p_D$.

@@

\begin{proof} The proof and the details of the lemma are attached to the supplementary documents. 
\end{proof}

@@

The conventional test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis, if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom.  It forms the benchmark upon which we evaluate the visual test. To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the conventional $p$-value as details given in Lemma \ref{lemma}. Thus the expected probability to reject by a single observer ($K=1$) in this scenario is

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$ gets larger, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. This dichotomy of visual power can also be seen in more detail in figure \ref{fig:power_loess_effect}.

@@

The experiment is designed to study the ability of human subjects to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated using a range of values of $\beta_2$ or slopes as shown in Table \ref{tbl:experiment_params}, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making  side-by-side boxplots of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and  plotted in the same way. The actual data plot was randomly placed among these null data plots to produce the lineup.  Figure \ref{fig:test_category} is an example of one of these lineups. It was generated for $n$=300, $\beta_2$=10 and $\sigma$=12. The actual data plot location is ($4^2-1$). For this lineup, 15 out of 16 observers picked the actual data plot.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the proportion correct ($\hat p$) is determined by the variance of the number of correct evaluations. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) we consider $\gamma =p$.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

@@

This experiment is very similar to the previous one, except that there is a single continuous covariate and no second covariate (Equation \ref{multi} with $p=1$), following the test in Table \ref{tbl:stat_multiple}, case 2. Data is simulated with two sample sizes ($n=100, 300$), two standard deviations of the error ($\sigma=5, 12$), a variety of slopes ($\beta$), as given in Table  \ref{tbl:experiment_params}. We arbitrarily fix $\beta_0 = 6$ and values for $X_1$ are simulated from $N(0,1)$. For each combination of parameters, at least three different actual data sets are produced, yielding a total of 70 lineups.
